---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r loading}
# reading data from file and getting summary
activity <- read.csv('activity.csv', colClasses = c('numeric', 'Date', 'numeric'))
summary(activity)
```
```{r processing}
# processing data and getting summary
library('stringr')
activity$interval <- str_pad(activity$interval, 4, pad='0')
activity$weekday <- as.factor(weekdays(activity$date))
summary(activity)
```

## What is mean total number of steps taken per day?
```{r reshaping data}
library('reshape2')
melted <- melt(activity, id=c('date'), measure.vars='steps', na.rm=TRUE)
result <- dcast(melted, date ~ variable, sum)
hist(result$steps, xlab='Number of steps', main='Histogram of the total number of steps taken each day')
mean(result$steps)
median(result$steps)
```

## What is the average daily activity pattern?
```{r getting daily activity pattern, fig.width=7}
library('reshape2')
melted <- melt(activity, id=c('interval'), measure.vars='steps', na.rm=TRUE)
result <- dcast(melted, interval ~ variable, mean)
summary(result)
with(result, plot(strptime(interval, '%H%M'), steps, type='l', xlab='Time', ylab='Number of steps in 5-minute interval', main='Average daily activity pattern'))
with(result, interval[which(steps == max(steps))])
```

## Imputing missing values
Number of missing values:
```{r calculation}
sum(is.na(activity$steps))
```
The strategy for filling gaps is the using median of step value for that interval for all days of observation.
```{r imputing}
clean_activity <- activity
empty_values_indexes <- which(is.na(activity$steps)) 
for (i in empty_values_indexes){
    interval_value <- activity$interval[i] 
    clean_activity$steps[i] <- with(result, steps[which(interval == interval_value)])
}
library('reshape2')
melted <- melt(clean_activity, id=c('date'), measure.vars='steps', na.rm=TRUE)
result <- dcast(melted, date ~ variable, sum)
hist(result$steps, xlab='Number of steps', main=expression(atop("Histogram of the total number of steps taken each day", " with removed NA's")))
```

## Are there differences in activity patterns between weekdays and weekends?
```{r, fig.width=14}
library('reshape2')
melted <- melt(subset(activity, !(weekday %in% c('Saturday', 'Sunday'))), id=c('interval'), measure.vars='steps', na.rm=TRUE)
weekdays <- dcast(melted, interval ~ variable, mean)
melted <- melt(subset(activity, weekday %in% c('Saturday', 'Sunday')), id=c('interval'), measure.vars='steps', na.rm=TRUE)
weekends <- dcast(melted, interval ~ variable, mean)
par(mfrow=c(1, 2))
with(weekdays, plot(strptime(interval, '%H%M'), steps, type='l', xlab='Time', ylab='Number of steps in 5-minute interval', main='Average weekdays activity pattern'))
with(weekends, plot(strptime(interval, '%H%M'), steps, type='l', xlab='Time', ylab='Number of steps in 5-minute interval', main='Average weekends activity pattern'))
#title('Differences in activity patterns between weekdays and weekends')
```